A White Noise Approach to Evolutionary Ecology

### 2020-04-11 .A White Noise Approach To Evolutionary Ecology/draft.Rmd
the models name is SAGA

Reviewing these accomplishments reveals a beautiful synthesis of evolution and population ecology. However, it also reveals a gap in theoretical approaches to incorporate the intrinsically random nature of populations. Specifically, in theoretical quantitative genetics the derivation of a population's response to random genetic drift is derived in discrete time under the assumption of constant effective population size using arguments based on properties of random samples [@Lande1976]. Though this approach conveniently mimics the formalism provided by the Wright-Fisher model of population genetics, real population sizes fluctuate over time. Furthermore, since these fluctuations are themselves stochastic, it seems natural to derive expressions for the evolutionary response to demographic stochasticity and consider how the results relate to characterizations of random genetic drift. This can be done in continuous time for population genetic models without too much technical overhead, assuming a finite number of alleles. However, for populations with a continuum of types, such as a quantitative trait, this becomes a vexing mathematical challenge. Here we close this gap by combining the calculus of white noise with results on rescaled limits of branching Brownian motion processes (BBM) and stochastic partial differential equations (SPDE). Our goal has two components: 1) Establish a novel synthetic approach to theoretical evolutionary ecology that provides a formal connection between demographic stochasticity and random genetic drift in the context of quantitative traits. To show that our approach can be used to develop useful biological insights we derive a model of coevolution in an ecological network and use it to investigate the relationship between competition coefficients and selection gradients. 2) Communicate some useful properties of space-time white noise, BBM and SPDE to as wide of audience as possible. With this goal in mind we will not provide a rigorous treatment of any of these deep subjects. Instead, we introduce a set of heuristics that only require the basic concepts of Reimann integration, partial differentiation and some exposure to Brownian motion and stochastic ordinary differential equations (SDE). For a concise introduction to SDE and Brownian motion, we recommend the primer by Evans [-@9781470410544]. Rigorous treatments of SPDE and rescaled limits of BBM can be found in Walsh [-@Walsh] and Dawson [-@dawson1993measure] respectively.

To provide motivation for the stochastic equations developed later and background for our model of coevolution, we begin with \S\ref{deterministic} by briefly summarizing derivations of deterministic dynamics of biological populations. Starting with a partial differential equation (PDE), we arrive at a general set of ordinary differential equations modelling the dynamics of abundance, trait mean and trait variance. From this we observe that replacing the PDE with a SPDE provides a path to derive SDE describing the evolutionary response to demographic stochasticity. We accomplish this in \S\ref{wnc} by introducing a set of mathematical tools based on the calculus of white noise and discuss how a diffusion limit of a spatially structured branching process leads to the natural SPDE appropriate for our study. The diffusion limit in turn provides a rigorous method for constructing fitness functions used in models of evolutionary ecology. We employ these tools to derive a system of SDE generalizing our deterministic results to account for demographic stochasticity. However, although biologically insightful, these equations remain difficult to analyze and implement numerically. In \S\ref{equations} we use an assumption of normally distributed trait values to simplify these expressions into formulae that are much easier to work with. We then account for the constraint of adaptive evolution on the amount of heritable variation in a population by extending these results via a model of imperfect inheritance. The resulting equations coincide with classical results in quantitative genetics as a special case. In \S\ref{coev} we combine the derived equations of population dynamics with classical niche theory to formulate a model of coevolution across a guild of $S$ species participating in exploitative competition along a common resource continuum. In SM \S\ref{diffuse} we apply a classical theorem on rescaled limits of BBM that allow for ecological interactions to provide a rigorous derivation. To gain biological insight, in \S\ref{dynamics} we numerically integrate our model of coevolution for $S=100$ species, tracking the dynamics of traits and abundances, under scenarios of weak and strong competition. We include an account of the natural history of the simulated system and discuss the significance of demographic stochasticity for structuring ecological communities. In \S\ref{ecoevo} we provide expressions for selection gradients and competition coefficients implied by our model and use these expressions to investigate the relationship between the degree of competition and coevolution. Finally, \S\ref{conclusion} concludes with a summary of accomplishments, a few remarks on the limits of this approach and future directions to incorporate more explicitly the genetic architecture of populations, feedbacks with ecosystem processes and the macroevolutionary history of interacting lineages.

@@

\begin{multline}\label{p_PDE}
\frac{\partial}{\partial t}p(x,t)=\frac{\partial}{\partial t}\frac{\nu(x,t)}{N(t)}=\frac{1}{N^2(t)}\left(N(t)\frac{\partial}{\partial t}\nu(x,t)-\nu(x,t)\frac{d}{dt}N(t)\right) \\
=\frac{1}{N(t)}\Big(m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)-\bar m(t)\nu(x,t)\Big) \\
=(m(\nu,x)-\bar m(t)) p(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}p(x,t).
\end{multline}
This result closely resembles Kimura's continuum-of-alleles model [@Kimura1965; @Brger1986]. The primary difference being that our model utilizes diffusion instead of convolution with an arbitrary mutation kernel. Of course, our model of mutation can be derived as an approximation to Kimura's model, which, for sexually reproducing populations, has been referred to as the Gaussian allelic approximation in reference to the distribution of mutational effects at loci in a genome on the values of traits [@Lande1975; but see also @9780471986539], the infinitesimal model in reference to modelling continuous traits as being encoded by an infinite number of loci each having infinitesimal effect [@Barton2017] and the Gaussian descendants approximation in reference to offspring trait values being normally distributed around their parental values [@Turelli2017]. Alternatively, since diffusion is the continuous-time equivalent to convolution against a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a special case of Kimura's continuum-of-alleles model. We return to these concepts in \S\ref{inheritance} where we review a simple model of inheritance that leads us to an expression for the stochastic evolution of additive genetic variance. To clarify the distinction between this model and previous models of phenotypic evolution we refer to it as the Deterministic Asexual Gaussian allelic model with Abundance dynamics (abbreviated DAGA). Later, we will extend this model to include the effects of demographic stochasticity, which we refer to as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA).

@@

In SM \S\ref{SDE_DERIV}, we extend these results to include the effects of demographic stochasticity. Similar to the approach taken by Champagnat, Ferriére and Méléard [-@Champagnat2006], we begin with a BBM that models populations as discrete sets of reproducing individuals whose vital rates depend on their trait value as well as the state of the entire population. Taking a large population size limit and keeping our assumption of single dimensional traits, we employ a pair of classical results that show, under the appropriate rescaling in time, phenotypic space and population density, a sequence of rescaled BBM converges to a limiting process that can be characterized by a SPDE [@Mlard1993; @zeng1998absolute]. The limiting processes of rescaled BBM have been referred to as measure-valued Markov processes [@dawson1993measure] or superprocesses [@alisonetheridge2000]. Under the simplifying assumptions inherited from our treatment of deterministic dynamics and the additional assumption that the magnitude of demographic stochasticity is independent of trait values, we obtain as a special case a relatively simple expression for an SPDE that generalizes equation (\ref{eq1}). The simplicity of our special case allows us to use properties of space-time white noise processes to derive a set of SDE that generalize equations (\ref{N_det}), (\ref{detmean}) and (\ref{vardyn_det}) to include the effects of demographic stochasticity. Classical expressions for the effects of random genetic drift on the evolution of mean traits are obtained as a special case.

@@

A convergence theorem for the diffusion limit of a generalization of BBM was established by Watanabe [-@Watanabe1968]. Dawson [-@Dawson1975] suggested that, for spatial dimension $d=1$, this diffusion limit should admit a density process that satisfies a SPDE. Konna and Shiga [-@Konno1988] and Reimers [-@Reimers1989] independently proved Dawson's suggestion was indeed correct. The diffusion limit of this more general branching process (in arbitrary spatial dimension) is referred to as a Dawson-Watanabe superprocess [@alisonetheridge2000]. Conditioning a Dawson-Watanabe superprocess to have constant mass returns a Fleming-Viot process [@Etheridge1991; @Perkins1991] which has been popular in studies of spatial population genetics. In particular, an extension of the Fleming-Viot process, known as the $\Lambda$-Fleming-Viot process, was introduced by Bertoin and Le Gall [-@Bertoin2003] and coined by Etheridge [-@Etheridge2008] where it was used to resolve some technical challenges in modelling isolation by distance [@Felsenstein1975; see also @Barton2013; and @Barton2019]. Although this provides an impressive list of accomplishments, the Dawson-Watanabe superprocess falls short of our needs. In particular this process assumes individuals do not interact and thus precludes its ability to model ecological interactions. However, this concern has been addressed, leading to constuctions of superprocesses that account for interactions among individuals. In the next subsection we summarize the main results in this area and introduce the SPDE that provides the basis for our approach to theoretical evolutionary ecology.

@@

The existence of diffusion limits for a class of measure-valued branching processes involving interactions among individuals has been treated by Méléard and Roelly [-@meleard1992interacting; -@Mlard1993]. The interactions can manifest as dependencies of the spatial movement or reproductive law of individuals on their position $x$ and the state of the whole population described by $X_t$. An important result of Méléard and Roelly [-@meleard1992interacting; -@Mlard1993] is a theorem that provides sufficient conditions to construct a sequence of rescaled measure-valued branching processes that converge to a generalization of the Dawson-Watanabe superprocess that includes interactions. The rescaling is analogous to that described above for non-interacting Dawson-Watanabe superprocesses, but now the reproductive law described by $\mathscr{W}(X_t,x)$ and $V(X_t,x)$, branching rate $\lambda(X_t,x)$ and diffusion parameter $\sqrt{\mu(X_t,x)}$ are allowed to depend on the whole population $X_t$ and individual location $x$. In Figure \ref{rescaled} we demonstrate this rescaling in discrete time for a population experiencing stabilizing selection and logistic growth. Since time is discretized, the process we simulate is formally a branching random walk. For further details on our simulation see SM \S\ref{numerical}.

@@

Our treatment of the relationship between breeding values and expressed traits follows classical quantitative genetic assumptions such as those used in the seminal paper by Lande [-@Lande1975] to investigate the maintenance of genetic variation. In particular, we assume that the expressed trait for any given individual is independent of environmental conditions and normally distributed around their breeding value with variance $\eta$. Hence, $\sigma^2=G+\eta$. In the case that all of the effects of alleles on an expressed trait are additive, $\eta$ is known as the _variance of environmental deviation_ [@Lande1975; @michaellynch1998]. For a given breeding value, we denote the probability density of a randomly drawn expressed trait by $\psi(x,g)$. Hence,
\begin{equation}
\psi(x,g)=\frac{1}{\sqrt{2\pi\eta}}\exp\left(-\frac{(x-g)^2}{2\eta}\right).
\end{equation}
To include this relationship in our framework, we write $\rho(g,t)$ as the abundance density of breeding values at time $t$ so that $\int_{-\infty}^{+\infty}\rho(g,t)dg=\int_{-\infty}^{+\infty}\nu(x,t)dx=N(t)$. We switch our focus from directly modelling the evolution of $\nu(x,t)$ to modelling the evolution of $\rho(g,t)$. Once $\rho(g,t)$ is determined, we can compute $\nu(x,t)$ via
\begin{equation}\label{nu_from_rho}
\nu(x,t)=\int_{-\infty}^{+\infty}\rho(g,t)\psi(x,g)dg.
\end{equation}
However, since selection acts on expressed phenotypes, we use the assumed relationship between breeding values and expressed traits to calculate the fitness of breeding values. To motivate the approach taken, consider the problem of inferring the breeding value of an individual given its expressed trait $x$. Denote $\mathfrak{g}$ a random variable representing the unkown breeding value. Under this model of inheritance we know $x$ is a random sample from a normal distribution with mean $\mathfrak{g}$ and variance $\eta$. Maximizing likelihood suggests $x$ is our best guess for $\mathfrak{g}$, but the actual value of $\mathfrak{g}$ is normally distributed around $x$ with the variance $\eta$. Hence, for fixed $x$, we obtain $\psi(x,g)$ as the probability density of $\mathfrak{g}$. Thus, the mean fitness of a breeding value $g$ across all individuals carrying $g$ can be written as
\begin{equation}\label{m_rel}
m^*(\rho,g)=\int_{-\infty}^{+\infty}m(\nu,x)\psi(x,g)dx.
\end{equation}
This is similar to the approach taken by Kimura and Crow [-@Kimura1978] to calculate the overall effects of selection for expressed characters onto the changes in the distribution of alleles encoding those characters. However, instead of focusing on the frequencies of alleles at particular loci, our results focus on the densities of breeding values. With the relationship between $m(\nu,x)$ and $m^*(\rho,g)$ established, we define the evolution of $\rho(g,t)$ by the SPDE
\begin{equation}\label{rho_SPDE}
\dot\rho(g,t)=m^*(\rho,g)\rho(g,t)+\frac{\mu}{2}\frac{\partial^2}{\partial^2 g}\rho(g,t)+\sqrt{V\rho(g,t)}\dot W(g,t).
\end{equation}
Equation (\ref{rho_SPDE}) is a stochastic generalization of the deterministic PDE (\ref{eq1}) from \S\ref{deterministic}, but describes the evolution of the distribution of breeding values instead of expressed characters. However, whether modelling expressed characters or breeding values, we refer to SPDE of the form (\ref{SPDE}) as Stochastic Asexual Gaussian allelic models with Abundance dynamics (abbreviated SAGA). By replacing $\psi(x,g)$ with different probability densities produces different developmental models. This includes densities that may depend on environmental variables and hence model phenotypic plasticy. To leave room for the naming of other models where $\psi(x,g)$ is Gaussian, we refer to the model of development we have chosen as the _centered_ Gaussian developmental model. In \S\ref{wnc} we review the origins of this equation and provide some theory to help make sense of it, particularly the term $\dot W$.

We assume $\rho(g,t)$ is Gaussian which implies its mode coincides with $\bar x$. Furthermore, since $\sigma^2=G+\eta$, we can use equation (\ref{m_rel}) and the chain rule from calculus (see SM \S\ref{fit2fit}) to find
\begin{subequations}
\begin{equation}
\frac{\partial\bar m}{\partial G}=\frac{\partial\bar m}{\partial\sigma^2}\frac{\partial\sigma^2}{\partial G}=\frac{\partial\bar m}{\partial\sigma^2},
\end{equation}
\begin{equation}
\overline{\frac{\partial m}{\partial G}}=\overline{\frac{\partial m}{\partial\sigma^2}\frac{\partial\sigma^2}{\partial G}}=\overline{\frac{\partial m}{\partial\sigma^2}}.
\end{equation}
\end{subequations}
Thus, equations (\ref{no_inher}) become
\begin{subequations}\label{inher}
\begin{equation}\label{xbarfinal}
d\bar x=G\left(\frac{\partial\bar m}{\partial\bar x}-\overline{\frac{\partial m}{\partial\bar x}}\right)dt+\sqrt{V\frac{G}{N}}dW_2,
\end{equation}
\begin{equation}\label{Gfinal}
dG=2G^2\left(\frac{\partial\bar m}{\partial G}-\overline{\frac{\partial m}{\partial G}}\right)dt+\left(\mu-V\frac{G}{N}\right)dt+ G\sqrt{\frac{2V}{N}}dW_3.
\end{equation}
\end{subequations}

@@

We have introduced an approach to derive models of evolutionary ecology using the calculus of white noise, developed SAGA, a novel model of phenotypic evolution that accounts for demographic stochasticity, demonstrated our approach by deriving a model of diffuse coevolution from classical competition theory, and used this model to investigate the relationship between selection gradients and competition coefficients. This work demonstrates that connecting contemporary theoretical approaches to evolutionary ecology with some fundamental results in the theory of measure-valued Markov processes allows for the development of a rigorous, yet flexible approach to synthesizing the dynamics of abundance and distributions of quantitative characters. Although this approach enjoys these important merits, there remains biological details and their associated technical challenges that need to be confronted for gaining a more thorough and accurate understanding of ecological communities. We touch on just four of them here and provide suggestions for future research to approach these challenges.

@@

Another approach to breaking the assumption of normally distributed trait values is the development of explicit multilocus models. These models describe the contributions of alleles at particular loci in the genome to the development of quantitative traits. Tracking the fluctuations of allele frequencies then allows theoreticians to investigate the dynamics of phenotypic distributions that deviate from normality. This approach has a long history in theoretical quantitative genetics [@bulmer1980; @pmid7851785; @Kirkpatrick1727] and coevolutionary theory [@Nuismer2005; @Kopp2006; @Nuismer2007]. However, work to investigate the role of complex genetic architectures in mediating feedbacks between the dynamics of population abundances and the distributions of traits mediating ecological interactions has apparently only just begun [@Patel2019].

>>>>

Reviewing these accomplishments reveals a beautiful synthesis of evolution and population ecology. However, it also reveals a gap in theoretical approaches to incorporate the intrinsically random nature of populations. Specifically, in theoretical quantitative genetics the derivation of a population's response to random genetic drift is derived in discrete time under the assumption of constant effective population size using arguments based on properties of random samples [@Lande1976]. Though this approach conveniently mimics the formalism provided by the Wright-Fisher model of population genetics, real population sizes fluctuate over time. Furthermore, since these fluctuations are themselves stochastic, it seems natural to derive expressions for the evolutionary response to demographic stochasticity and consider how the results relate to characterizations of random genetic drift. This can be done in continuous time for population genetic models without too much technical overhead, assuming a finite number of alleles. However, for populations with a continuum of types, such as a quantitative trait, this becomes a vexing mathematical challenge. Here we close this gap by combining the calculus of white noise with results on rescaled limits of branching Brownian motion processes (BBM) and stochastic partial differential equations (SPDE), which are stochastic analogs of deterministic partial differential equations (PDE). Our goal has two components: 1) Establish a novel synthetic approach to theoretical evolutionary ecology that provides a formal connection between demographic stochasticity and random genetic drift in the context of quantitative traits. To show that our approach can be used to develop useful biological insights we derive a model of coevolution in an ecological network and use it to investigate the relationship between competition coefficients and selection gradients. 2) Communicate some useful properties of space-time white noise, BBM and SPDE to as wide of audience as possible. With this goal in mind we will not provide a rigorous treatment of any of these deep subjects. Instead, we introduce a set of heuristics that only require the basic concepts of Reimann integration, partial differentiation and some exposure to Brownian motion and stochastic ordinary differential equations (SDE). For a concise introduction to SDE and Brownian motion, we recommend the primer by Evans [-@9781470410544]. Rigorous treatments of SPDE and rescaled limits of BBM can be found in Walsh [-@Walsh] and Dawson [-@dawson1993measure] respectively.

To provide motivation for the stochastic equations developed later and background for our model of coevolution, we begin with \S\ref{deterministic} by briefly summarizing derivations of deterministic dynamics of biological populations. Starting with a PDE (which we refer to as the Deterministic Asexual Gaussian allelic model with Abundance dynamics, abbreviated DAGA), we arrive at a general set of ordinary differential equations modelling the dynamics of abundance, trait mean and trait variance. We discuss the relationship between DAGA and previous models of phenotypic evolution that have been employed in theoretic quantitative genetics. Observing that generalizing the PDE to a SPDE, we find a path to derive SDE describing the evolutionary response to demographic stochasticity. We accomplish this in \S\ref{wnc} by introducing a set of mathematical tools based on the calculus of white noise and discuss how a diffusion limit of a spatially structured branching process leads to the natural SPDE appropriate for our study. We refer to this SPDE as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA). The diffusion limit in turn provides a rigorous method for constructing fitness functions used in models of evolutionary ecology. We employ these tools to derive a system of SDE generalizing our deterministic results to account for demographic stochasticity. We find that the effect of random genetic drift on the evolution of mean trait and phenotypic variance characterizes the process of random genetic drift in continuous time. However, although biologically insightful, these equations remain difficult to analyze and implement numerically. In \S\ref{equations} we use an assumption of normally distributed trait values to simplify these expressions into formulae that are much easier to work with. We then account for the constraint of adaptive evolution on the amount of heritable variation in a population by extending these results via a model of imperfect inheritance. The resulting equations coincide with classical results in quantitative genetics as a special case. In \S\ref{coev} we combine the derived equations of population dynamics with classical niche theory to formulate a model of coevolution across a guild of $S$ species participating in exploitative competition along a common resource continuum. In SM \S\ref{diffuse} we apply a classical theorem on rescaled limits of BBM that allow for ecological interactions to provide a rigorous derivation. To gain biological insight, in \S\ref{dynamics} we numerically integrate our model of coevolution for $S=100$ species, tracking the dynamics of traits and abundances under scenarios of weak and strong competition. We include an account of the natural history of the simulated system and discuss the significance of demographic stochasticity for structuring ecological communities. In \S\ref{ecoevo} we provide expressions for selection gradients and competition coefficients implied by our model and use these expressions to investigate the relationship between the degree of competition and coevolution. Finally, \S\ref{conclusion} concludes with a summary of accomplishments, a few remarks on the limits of this approach and future directions to incorporate more explicitly the effects of small populations, the genetic architecture of quantitative traits, feedbacks with ecosystem processes and the macroevolutionary history of interacting lineages.

@@

\begin{multline}\label{p_PDE}
\frac{\partial}{\partial t}p(x,t)=\frac{\partial}{\partial t}\frac{\nu(x,t)}{N(t)}=\frac{1}{N^2(t)}\left(N(t)\frac{\partial}{\partial t}\nu(x,t)-\nu(x,t)\frac{d}{dt}N(t)\right) \\
=\frac{1}{N(t)}\Big(m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)-\bar m(t)\nu(x,t)\Big) \\
=(m(\nu,x)-\bar m(t)) p(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}p(x,t).
\end{multline}
This result closely resembles Kimura's continuum-of-alleles model [@Kimura1965; @Brger1986]. The primary difference being that our model utilizes diffusion instead of convolution with an arbitrary mutation kernel. Of course, our model of mutation can be derived as an approximation to Kimura's model, which, for sexually reproducing populations, has been referred to as the Gaussian allelic approximation in reference to the distribution of mutational effects at loci in a genome on the values of traits [@Lande1975; but see also @9780471986539], the infinitesimal model in reference to modelling continuous traits as being encoded by an infinite number of loci each having infinitesimal effect [@Barton2017] and the Gaussian descendants approximation in reference to offspring trait values being normally distributed around their parental values [@Turelli2017]. Alternatively, since diffusion is the continuous-time equivalent to convolution against a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a special case of Kimura's continuum-of-alleles model. We return to these concepts in \S\ref{inheritance} where we review a simple model of inheritance that leads us to an expression for the stochastic evolution of additive genetic variance. To clarify the distinction between this model and previous models of phenotypic evolution we refer to the PDE (\ref{eq1}) from which (\ref{p_PDE}) was derived as the Deterministic Asexual Gaussian allelic model with Abundance dynamics (abbreviated DAGA). Later, we will extend this model to include the effects of demographic stochasticity, which we refer to as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA).

@@

In SM \S\ref{SDE_DERIV}, we extend these results to include the effects of demographic stochasticity. The key realization in this direction is that a rigorous derivation of a SPDE that naturally generalizes (\ref{eq1}) in the context of biological populations that accounts for ecological interactions has been provided by Li [-@zeng1998absolute]. To understand how this SPDE was derived, we provide an informal discussion of measure-valued branching processes and their diffusion limits. Similar to the approach taken by Champagnat, Ferriére and Méléard [-@Champagnat2006], we begin with a BBM that models populations as discrete sets of reproducing individuals whose vital rates depend on their trait value as well as the state of the entire population. Taking a large population size limit and keeping our assumption of single dimensional traits, we employ a pair of classical results that show, under the appropriate rescaling in time, phenotypic space and population density, a sequence of rescaled BBM converges to a limiting process that can be characterized by a SPDE [@Mlard1993; @zeng1998absolute]. The limiting processes of rescaled BBM have been referred to as measure-valued Markov processes [@dawson1993measure] or superprocesses [@alisonetheridge2000]. Under the simplifying assumptions inherited from our treatment of deterministic dynamics and the additional assumption that the magnitude of demographic stochasticity is independent of trait values, we obtain as a special case a relatively simple expression for an SPDE that generalizes equation (\ref{eq1}). The simplicity of our special case allows us to use properties of space-time white noise processes to derive a set of SDE that generalize equations (\ref{N_det}), (\ref{detmean}) and (\ref{vardyn_det}) to include the effects of demographic stochasticity. Classical expressions for the effects of random genetic drift on the evolution of mean traits are obtained as a special case.

@@

A convergence theorem for the diffusion limit of a generalization of BBM was established by Watanabe [-@Watanabe1968]. Dawson [-@Dawson1975] suggested that, for spatial dimension $d=1$, this diffusion limit should admit a density process that satisfies a SPDE. Konna and Shiga [-@Konno1988] and Reimers [-@Reimers1989] independently proved Dawson's suggestion was indeed correct. The diffusion limit of this more general branching process (in arbitrary spatial dimension) is referred to as a Dawson-Watanabe superprocess [@alisonetheridge2000]. Conditioning a Dawson-Watanabe superprocess to have constant mass returns a Fleming-Viot process [@Etheridge1991; @Perkins1991] which has been popular in studies of spatial population genetics. In particular, an extension of the Fleming-Viot process, known as the $\Lambda$-Fleming-Viot process, was introduced by Bertoin and Le Gall [-@Bertoin2003] and coined by Etheridge [-@Etheridge2008] where it was used to resolve some technical challenges in modelling isolation by distance [@Felsenstein1975; see also @Barton2013; and @Barton2019]. Although this provides an impressive list of accomplishments, the Dawson-Watanabe superprocess falls short of our needs. In particular this process assumes individuals do not interact and thus precludes its ability to model nonlocal effects on the fitness of individuals, such as those produced via ecological interactions. However, this concern has been addressed, leading to constuctions of superprocesses that account for interactions among individuals. In the next subsection we summarize the main results in this area and introduce the SPDE that provides the basis for our approach to theoretical evolutionary ecology.

@@

In the above subsection we reviewed the diffusion limit of an especially tractable measure-valued branching process. However, we found the simplicity of this process restricts us from modelling nonlocal effects on the fitness of individuals. Here, we discuss superprocesses that account for such effects. The existence of diffusion limits for a class of measure-valued branching processes involving nonlocal effects has been treated by Méléard and Roelly [-@meleard1992interacting; -@Mlard1993]. The interactions can manifest as dependencies of the spatial movement or reproductive law of individuals on their position $x$ and the state of the whole population described by $X_t$. An important result of Méléard and Roelly [-@meleard1992interacting; -@Mlard1993] is a theorem that provides sufficient conditions to construct a sequence of rescaled measure-valued branching processes that converge to a generalization of the Dawson-Watanabe superprocess that includes interactions. The rescaling is analogous to that described above for non-interacting Dawson-Watanabe superprocesses, but now the reproductive law described by $\mathscr{W}(X_t,x)$ and $V(X_t,x)$, branching rate $\lambda(X_t,x)$ and diffusion parameter $\sqrt{\mu(X_t,x)}$ are allowed to depend on the whole population $X_t$ and individual location $x$. In Figure \ref{rescaled} we demonstrate this rescaling in discrete time for a population experiencing stabilizing selection and logistic growth. Since time is discretized, the process we simulate is formally a branching random walk. For further details on our simulation see SM \S\ref{numerical}.

@@

Our treatment of the relationship between breeding values and expressed traits follows classical quantitative genetic assumptions such as those used in the seminal paper by Lande [-@Lande1975] to investigate the maintenance of genetic variation. In particular, we assume that the expressed trait for any given individual is independent of environmental conditions and normally distributed around their breeding value with variance $\eta$. Hence, $\sigma^2=G+\eta$. In the case that all of the effects of alleles on an expressed trait are additive, $\eta$ is known as the _variance of environmental deviation_ [@Lande1975; @michaellynch1998]. For a given breeding value, we denote the probability density of a randomly drawn expressed trait by $\psi(x,g)$. Hence,
\begin{equation}
\psi(x,g)=\frac{1}{\sqrt{2\pi\eta}}\exp\left(-\frac{(x-g)^2}{2\eta}\right).
\end{equation}
To include this relationship in our framework, we write $\rho(g,t)$ as the abundance density of breeding values at time $t$ so that $\int_{-\infty}^{+\infty}\rho(g,t)dg=\int_{-\infty}^{+\infty}\nu(x,t)dx=N(t)$. We switch our focus from directly modelling the evolution of $\nu(x,t)$ to modelling the evolution of $\rho(g,t)$. Once $\rho(g,t)$ is determined, we can compute $\nu(x,t)$ via
\begin{equation}\label{nu_from_rho}
\nu(x,t)=\int_{-\infty}^{+\infty}\rho(g,t)\psi(x,g)dg.
\end{equation}
However, since selection acts on expressed phenotypes, we use the assumed relationship between breeding values and expressed traits to calculate the fitness of breeding values. To motivate the approach taken, consider the problem of inferring the breeding value of an individual given its expressed trait $x$. Denote $\mathfrak{g}$ a random variable representing the unkown breeding value. Under this model of inheritance we know $x$ is a random sample from a normal distribution with mean $\mathfrak{g}$ and variance $\eta$. Maximizing likelihood suggests $x$ is our best guess for $\mathfrak{g}$, but the actual value of $\mathfrak{g}$ is normally distributed around $x$ with the variance $\eta$. Hence, for fixed $x$, we obtain $\psi(x,g)$ as the probability density of $\mathfrak{g}$. Thus, the mean fitness of a breeding value $g$ across all individuals carrying $g$ can be written as
\begin{equation}\label{m_rel}
m^*(\rho,g)=\int_{-\infty}^{+\infty}m(\nu,x)\psi(x,g)dx.
\end{equation}
This is similar to the approach taken by Kimura and Crow [-@Kimura1978] to calculate the overall effects of selection for expressed characters onto the changes in the distribution of alleles encoding those characters. However, instead of focusing on the frequencies of alleles at particular loci, our results focus on the densities of breeding values. With the relationship between $m(\nu,x)$ and $m^*(\rho,g)$ established, we define the evolution of $\rho(g,t)$ by the SPDE
\begin{equation}\label{rho_SPDE}
\dot\rho(g,t)=m^*(\rho,g)\rho(g,t)+\frac{\mu}{2}\frac{\partial^2}{\partial^2 g}\rho(g,t)+\sqrt{V\rho(g,t)}\dot W(g,t).
\end{equation}
Equation (\ref{rho_SPDE}) is a stochastic generalization of the deterministic PDE (\ref{eq1}) from \S\ref{deterministic}, but describes the evolution of the distribution of breeding values instead of expressed characters. However, whether modelling expressed characters or breeding values, we refer to SPDE of the form (\ref{rho_SPDE}) as Stochastic Asexual Gaussian allelic models with Abundance dynamics (abbreviated SAGA). By replacing $\psi(x,g)$ with different probability densities produces different developmental models. This includes densities that may depend on environmental variables and hence model phenotypic plasticy. To leave room for the naming of other models where $\psi(x,g)$ is Gaussian, we refer to the model of development we have chosen as the _centered_ Gaussian developmental model. In \S\ref{wnc} we review the origins of equation (\ref{rho_SPDE}) and provide some theory to help make sense of it, particularly the term $\dot W$.

Assuming $\rho(g,t)$ is Gaussian implies its mode coincides with $\bar x$. Furthermore, since $\sigma^2=G+\eta$, we can use equation (\ref{m_rel}) and the chain rule from calculus (see SM \S\ref{fit2fit}) to find
\begin{subequations}
\begin{equation}
\frac{\partial\bar m}{\partial G}=\frac{\partial\bar m}{\partial\sigma^2}\frac{\partial\sigma^2}{\partial G}=\frac{\partial\bar m}{\partial\sigma^2},
\end{equation}
\begin{equation}
\overline{\frac{\partial m}{\partial G}}=\overline{\frac{\partial m}{\partial\sigma^2}\frac{\partial\sigma^2}{\partial G}}=\overline{\frac{\partial m}{\partial\sigma^2}}.
\end{equation}
\end{subequations}
Thus, equations (\ref{no_inher}) become
\begin{subequations}\label{inher}
\begin{equation}\label{xbarfinal}
d\bar x=G\left(\frac{\partial\bar m}{\partial\bar x}-\overline{\frac{\partial m}{\partial\bar x}}\right)dt+\sqrt{V\frac{G}{N}}dW_2,
\end{equation}
\begin{equation}\label{Gfinal}
dG=2G^2\left(\frac{\partial\bar m}{\partial G}-\overline{\frac{\partial m}{\partial G}}\right)dt+\left(\mu-V\frac{G}{N}\right)dt+ G\sqrt{\frac{2V}{N}}dW_3.
\end{equation}
\end{subequations}

@@

We have introduced an approach to derive models of evolutionary ecology using the calculus of white noise and developed SAGA, a novel model of phenotypic evolution that accounts for demographic stochasticity. From SAGA we derived SDE that track the evolution of abundance, mean trait and additive genetic variance. Observing the expressions of these SDE we find the effects of demographic stochasticity on the evolution of mean trait and additive genetic variance characterize the effects of random genetic drift. By considering the statistics of random samples, this has previously been in models of fixed population size in discrete time [@Lande1976]. However, we have shown random genetic drift is a result of demographic stochasticity for populations with fluctuating abundances in continuous time. To demonstrate the relevance of our approach to studies of evolutionary ecology, we comined our SDE with classical competition theory to derive a model of diffuse coevolution and used this model to investigate the relationship between selection gradients and competition coefficients. This work demonstrates that connecting contemporary theoretical approaches to evolutionary ecology with some fundamental results in the theory of measure-valued Markov processes allows for the development of a rigorous, yet flexible approach to synthesizing the dynamics of abundance and distributions of quantitative characters, in agreement with Champagnat et. al. [-@Champagnat2006]. Although this approach enjoys these important merits, there remains biological details and their associated technical challenges that need to be confronted for gaining a more thorough and accurate understanding of ecological communities. We touch on just four of them here and provide suggestions for future research to approach these challenges. For the first three, we recommend the use of individual-based models such as BBM to confront these challenges. To summarize why, we state here that analytical results become difficult to derive for small populations and complicated models of mutation and selection.

@@

Another approach to breaking the assumption of normally distributed trait values is the development of explicit multilocus models. These models describe the contributions of alleles at particular loci in the genome to the development of quantitative traits. Tracking the fluctuations of allele frequencies then allows theoreticians to investigate the dynamics of phenotypic distributions that deviate from normality. This approach has a long history in theoretical quantitative genetics [@bulmer1980; @pmid7851785; @Kirkpatrick1727] and coevolutionary theory [@Nuismer2005; @Kopp2006; @Nuismer2007]. However, work to investigate the role of complex genetic architectures in mediating feedbacks between the dynamics of population abundances and the distributions of traits mediating ecological interactions has apparently only just begun [@Patel2019]. However, since analytical results are difficult to obtain for general models of mutation, development and selection  [but see @9780471986539], we again suggest the use of individual-based simulations. For example, by assuming the mutational effects in the BBM model is Gaussian with variance an order of magnitude greater than the equilibrium phenotypic (or genetic) variance, we obtain the so-called _House-of-cards_ model [@Turelli1984]. However, perhaps considering heavy-tailed distributions, such as the Laplace-distributions considered in Park et. al. [-@Park2017], will provide more realistic models of mutational effects. This has important implications in conservation genetics where small populations may evolve primarily in response to mutation and random genetic drift [@Lande1995].
